Tsbrowse: a genome browser for Ancestral Recombination Graphs

### 2025-03-27 manuscript/paper.tex
Update to discussion

Interest in ARG-based methods is burgeoning, but the methods are new 
and practical guidance on applying inferences to real data is lacking.
Data filtering is essential, and the effects of the 
choices that must be made along any bioinformatics pipeline on the final
ARG are hard to predict and quantify. \texttt{Tsbrowse} was primarily
developed as a way to quickly visualise the effects of such filtering
choices on ARGs inferred by \texttt{tsinfer}, and it is now an indispensible
element of the inference pipeline. Fig~\ref{fig:Supplementary_Figure_2} 
shows [a before and after view for 1000G data, where tsbrowse helped 
diagnose and resolve an opaque data problem].
These interactive visualisations have also helped diagnose issues with
the \texttt{tsinfer} inference algorithm. 
Fig~\ref{fig:Supplementary_Figure_3} shows the genomic spans of 
ancestral nodes in an ARG inferred with tsinfer, demonstrating a clear 
excess of long haplotypes in the very ancient past. These insights 
have helped guide development and 
may lead to significant improvements in performance.

@@

\section{Discussion} \label{sec:Discussion} Visualisation techniques are
essential for data-driven insights. Specialised tools such as \textit{Treenome
browser} \citep{Kramer2023} tackle the challenge of semantic visualisation of
large phylogenies. Yet capturing the data model underlying ARGs is crucial for
the systematic exploration of local genealogies. \texttt{Tsbrowse} addresses
this open problem by condensing ARG properties into a set of browsable,
interactive visual summaries, thus aiding in ARG-based data exploration and
hypothesis generation.

>>>>

Interest in ARG-based methods is burgeoning, but the methods are new 
and practical guidance on applying inferences to real data is lacking.
Data filtering is essential, and the effects of the 
choices that must be made along any bioinformatics pipeline on the final
ARG are hard to predict and quantify. \texttt{Tsbrowse} was primarily
developed as a way to quickly visualise the effects of such filtering
choices on ARGs inferred by \texttt{tsinfer}, and it is now an indispensible
element of the inference pipeline. Fig~\ref{fig:Supplementary_Figure_2} 
shows [a before and after view for 1000G data, where tsbrowse helped 
diagnose and resolve an opaque data problem].
These interactive visualisations have also helped diagnose issues with
the \texttt{tsinfer} inference algorithm. 
Fig~\ref{fig:Supplementary_Figure_3} shows the genomic spans of 
ancestral nodes in an ARG inferred with \texttt{tsinfer}, demonstrating a clear 
excess of long haplotypes in the very ancient past. These insights 
have helped guide development and 
may lead to significant improvements in performance.

@@

\section{Discussion} \label{sec:Discussion} 
Visualisation of tree topology is a central task in phylogenetics, 
and although numerous tools
exist~\citep[e.g.][]{huson2007dendroscope,vaughan2017icytree},
the methods can typically only handle a few hundred nodes.
Visualisation of large-scale tree topologies with millions of nodes
requires much more sophisticated approaches
to capture topological features at different scales, 
and is active research area~\citep{wong2022dynamic,kramer2023treenome}.
Adapting such methods, and integrating into \texttt{tsbrowse} to provide
a local tree viewer that operates at the million-node scale
is an important direction for future work.
An interactive viewer for the entire ARG topology, capturing 
semantic properties of the graph at a range of scales, is an even more 
ambitious goal, and would be a major asset for the field.
